A fully labelled proof system for intuitionistic modal logics

### 2020-09-22 FullyLabelled/JLC/review-answer.txt

We never claim the birelational semantics to be a novelty (it goes
back to the 1980s). Our paper is not about semantics but about about
proof systems. And our system allows for extenstions that could not be
studied before. We made this more clear in the new version (see
Section 7).

>  Also the reason for preferring one semantics for intuitionistic
>  modal logic over another should be discussed in detail, as
>  otherwise the authors’ work would rather look a routine exercise

We are puzzled by this comment, as we just picked the standard
semantics for intuitionistic modal logic, and semantics is not a
concern of our paper. However, we added in the conclusion some
comments on constructive modal logics, which are another variant of
adding BOX and DIAMOND to intuitionistic propositional logic.

Reviewer 2 finds this paper well written, but has some technical
concerns. We address them below.

>  The main issue is indeed with the "wide range of logics”. While the
>  base logic is discussed in detail, the range of logics is treated
>  in an “Extensions” section that I believe could actually be more
>  detailed.

@@

TODO

>  Most importantly, I am not convinced by the gklmn rule. The side
>  condition stipulates that y’ and u are fresh, whereas the frame
>  property contains existential quantifications. I was thus expecting
>  either a Skolemization in the style of Vigano in the “Labelled
>  Non-Classical Logics” book, or a geometric rule with existential
>  quantification in the style of Simpson [Sim94] or, perhaps even
>  better, a “mixed” rule where the labels are assumed more or less in
>  the style of an \exists elimination to derive a labeled formula.  I
>  think that a thorough discussion of the alternatives, along with a
>  more thorough proof of the correctness of the chosen solution, is
>  needed.

We added Remark 7.5 to discuss this. Our rules are in fact in the
style of Simpson, but we added a comparison to Vigano's method.

>  By the way, it does not help at all that the applications of the
>  rule g1111 are merged with \Box_L and \Diamond_R in the example
>  derivation.

We fixed that.

>  Also Remark 4.2 could benefit from some additional explanation. You
>  write that the proof of Theorem 4.1 shows the “need” of the rules
>  F_1, refl and trans. However, strictly speaking, the proof only
>  shows how the rules are used and that they are helpful to prove the
>  axioms. To show the “need" you would likely have to show that no
>  other proofs of the axioms are possible. That is of course the
>  case, but unless you prove it formally, it is not fully correct to
>  claim that the “need” is shown. This is pedantic, I know.

We just removed the "need".

### 2020-09-28 FullyLabelled/JLC/review-answer.txt

Dear Didier,

thank you very much for the reviews. We are very grateful to the
reviewers for their comments, which helped us improve our paper.

@@

We are puzzled by this comment, as we just picked the standard
semantics for intuitionistic modal logic, and semantics is not a
concern of our paper. However, we added in the conclusion some
comments on constructive modal logics, which are another variant of
adding BOX and DIAMOND to intuitionistic propositional logic.

Reviewer 2 finds this paper well written, but has some technical
concerns. We address them below.

>  The main issue is indeed with the "wide range of logics”. While the
>  base logic is discussed in detail, the range of logics is treated
>  in an “Extensions” section that I believe could actually be more
>  detailed.

We extended this section with more explanation and also added a few
more comments and remarks.

### 2020-09-28 FullyLabelled/JLC/review-answer.txt

Dear Didier,

thank you very much for the reviews. We are very grateful to the
reviewers for their comments, which helped us improve our paper.

We took all comments and remarks into account, and below we address
the reviewers comments in detail.

One of the main concerns of Reviewer 1 is that we have not cited all
the relevant literature. We thank the reviewer for pointing out the
missing references and we added them accordingly.

@@

We now address the other comments:

>  Perhaps even more importantly, the authors of the paper under
>  review neglect an important caveat in the design of rules of
>  contraction-free labelled sequent calculi: to guarantee that
>  contraction be admissible one needs to ensure that a certain
>  closure condition be satisfied

We already said in the submitted version that in our system
contraction is derivable from the monotonicity rules, which in turn
are admissible. We made this even more clear in the new version in
Remark 3.4.

@@

>  In Section 7, if the rules G_klmn are added without their
>  contracted instances, the calculus will not be complete. For
>  example, in the classical case, Euclidean transitivity ∀a, b,
>  c(aRb&aRc → bRc) has the contracted instance ∀a, b(aRb → bRb) that
>  allows for the derivation of BOX(BOX A → A). This, however, can
>  hardly be derived without the rule’s contracted instance.

In fact, the contracted instances are added. In the same way as for
other rules, occurring labels can be equal. We failed to make this
explicit in the first submitted version and corrected that mistake in Remark 7.2.

@@

We never claim the birelational semantics to be a novelty (it goes
back to the 1980s). Our paper is not about semantics but about
proof systems. And our system allows for extenstions that could not be 
studied before. We made this more clear in the new version (see
Section 7).

>  Also the reason for preferring one semantics for intuitionistic
>  modal logic over another should be discussed in detail, as
>  otherwise the authors’ work would rather look a routine exercise

We are puzzled by this comment, as we just picked the standard
semantics for intuitionistic modal logic, and semantics is not a
concern of our paper. However, we added in the conclusion some
comments on constructive modal logics, which are another variant of
adding BOX and DIAMOND to intuitionistic propositional logic.

On the other side, Reviewer 2 finds this paper well written, but has some 
technical concerns. We address them below.

@@

>  The main issue is indeed with the "wide range of logics”. While the
>  base logic is discussed in detail, the range of logics is treated
>  in an “Extensions” section that I believe could actually be more
>  detailed.

We put more emphasis on the section 7 in order to make clearer how to generate "stronger
logics" adding new axioms to our system. We extended this section with more explanation and 
we also added a few more comments and remarks. For example, as a better way to understand the extensions, 
we also illustrated the specific case of adding to IK the 4-axiom.

@@

>  Most importantly, I am not convinced by the gklmn rule. The side
>  condition stipulates that y’ and u are fresh, whereas the frame
>  property contains existential quantifications. I was thus expecting
>  either a Skolemization in the style of Vigano in the “Labelled
>  Non-Classical Logics” book, or a geometric rule with existential
>  quantification in the style of Simpson [Sim94] or, perhaps even
>  better, a “mixed” rule where the labels are assumed more or less in
>  the style of an \exists elimination to derive a labeled formula.  I
>  think that a thorough discussion of the alternatives, along with a
>  more thorough proof of the correctness of the chosen solution, is
>  needed.

\marianela{not sure what to put here} We added Remark 7.5 to discuss this. 
Our rules are in fact in the style of Simpson, but we added a comparison to Vigano's method.

@@

>  By the way, it does not help at all that the applications of the
>  rule g1111 are merged with \Box_L and \Diamond_R in the example
>  derivation.

Following this comment, we fixed the derivation on the paper in order to 
show better the use of the rule g1111.

@@

>  Also Remark 4.2 could benefit from some additional explanation. You
>  write that the proof of Theorem 4.1 shows the “need” of the rules
>  F_1, refl and trans. However, strictly speaking, the proof only
>  shows how the rules are used and that they are helpful to prove the
>  axioms. To show the “need" you would likely have to show that no
>  other proofs of the axioms are possible. That is of course the
>  case, but unless you prove it formally, it is not fully correct to
>  claim that the “need” is shown. This is pedantic, I know.

In this case, we chose to change the word "need" with the word "use".
Our intention on this, was to try to show how to apply those rules and how useful there are
in our proofs.

### 2020-09-28 FullyLabelled/JLC/review-answer.txt

We are puzzled by this comment, as we just picked the standard
semantics for intuitionistic modal logic, and semantics is not a
concern of our paper. However, we added in the conclusion some
comments on constructive modal logics, which are another variant of
adding BOX and DIAMOND to intuitionistic propositional logic.

On the other side, Reviewer 2 finds this paper well written, but has some 
technical concerns. We address them below.

>  The main issue is indeed with the "wide range of logics”. While the
>  base logic is discussed in detail, the range of logics is treated
>  in an “Extensions” section that I believe could actually be more
>  detailed.

We put more emphasis on the section 7 in order to make more clear how to 
generate "stronger logics" adding new axioms to our system. We extended 
this section with more explanation and we also added a few more comments 
and remarks. For example, as a better way to understand the extensions, 
we also illustrated the specific case of adding to IK the 4-axiom.

@@

Following this comment, we fixed the derivation on the paper in order to 
show better the use of the rule g1111.

>  Also Remark 4.2 could benefit from some additional explanation. You
>  write that the proof of Theorem 4.1 shows the “need” of the rules
>  F_1, refl and trans. However, strictly speaking, the proof only
>  shows how the rules are used and that they are helpful to prove the
>  axioms. To show the “need" you would likely have to show that no
>  other proofs of the axioms are possible. That is of course the
>  case, but unless you prove it formally, it is not fully correct to
>  claim that the “need” is shown. This is pedantic, I know.

In this case, we chose to change the word "need" with the word "use".
Our intention on this, was to try to show how to apply those rules and how 
useful there are in our proofs.

>Reconsider de abstract. It reads very operational and does not clearly state the
>contributions of the paper.

We also consider this comment of the Reviewer with the aim of better reflecting 
our contribution in the paper.

Finally, as the Reviwer 2 pointed out, we fixed some broken references.

We want to thank the reviewers again for their time and for helping us improve our paper. 
We hope to be able to show our interest in answering each of the points marked by each reviewer.

### 2020-09-29 FullyLabelled/JLC/conclusion.tex
progress on section 7

>>>>

\review{Most importantly, I am not convinced by the gklmn rule. The side condition stipulates that $\lb y'$ and $\lb u$ are fresh, whereas the frame property contains existential quantifications. I was thus expecting either a Skolemization in the style of Vigano in the ``Labelled Non-Classical Logics" book, or a geometric rule with existential quantification in the style of Simpson [Sim94] or, perhaps even better, a “mixed” rule where the labels are	assumed more or less in the style of an $\exists$ elimination to derive a labeled formula.}
\todo{add coment on this}

The rules $gklmn$ (similarly to those corresponding to reflexivity, transitivity, $F_1$ and $F_2$) are obtained through the standard axioms-to-rule procedure that is well known for geometric axioms, and applies more generally to bipolar axioms~\cite{marin:etal:submitted} (by assuming a positive bias on atoms).
%
On the other hand, a different shape of rules can be obtained in the style of Vigan\`o~\cite{vigano:00 } (by assuming atoms are negative polarised). 
%
For reflexivity and transitivity of the preorder relation, this would give:
%
$$
\vlinf{}{}{\Gamma \SEQ \Delta, \futs xx}{}
\qquad
\vliinf{}{}{\Gamma \SEQ \Delta, \futs xz}{\Gamma \SEQ \Delta, \futs xy}{\Gamma \SEQ \Delta, \futs yz}
$$
%
We conjecture that $\ax{F_1}$ could be incorporated to the system as a pair of rules of the following form:
$$\vliinf{}{}{\B, \Gamma \SEQ \Delta, \futs x{f(x,y,z)}}{\B, \Gamma \SEQ \Delta, \accs xy}{\B, \Gamma \SEQ \Delta, \futs yz}
\qquad
\vliinf{}{}{\B, \Gamma \SEQ \Delta, \accs {g(x,y,z)}z}{\B, \Gamma \SEQ \Delta, \accs xy}{\B, \Gamma \SEQ \Delta, \futs yz}$$
%
where $f$ and $g$ are (Skolem) function constants.
%
Note however that these require (relational and preorder) atoms to appear on the right-hand-side of the sequent and are to be read top-down, and therefore are not appropriate for proof search.	
%	\lutz{we should say that this is pretty pointless wrt proof search.}
%
It should also possible to design a ``mixed" rule, by considering relational atoms positively and pre-order atoms negatively, or vice-versa, and we leave this investigation for future work.
%$$
%\vliinf{}{\lb u \mbox{ fresh}}{\B, \accs xy, \Gamma \SEQ \Delta}{\B, \accs xy, \futs yz, \Gamma \SEQ \Delta}{\accs xy, \futs xu, \accs uz, \Gamma \SEQ \Delta}
%\qquad	
%\vliinf{}{\lb u \mbox{ fresh}}{\B, \futs yz, \Gamma \SEQ \Delta}{\B, \futs yz, \accs xy, \Gamma \SEQ \Delta}{\futs yz, \futs xu, \accs uz, \Gamma \SEQ \Delta}$$

### 2020-09-29 FullyLabelled/JLC/introduction.tex
progress on section 7

Standard labelled sequent calculi attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally use
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame conditions corresponding to the desired logic can be directly
encoded as inference rules.
\todo{Reviewer 1 complained about the ``best''}
Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
\label{eq:Rtrans}
\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

@@

	logics}, whose Kripke semantics is based on \emph{birelational}

@@

	\fm{\BOX A\IMP \BOX\BOX A}

@@

	\fm{\DIA\DIA A\IMP\DIA A}
	\quadcm

@@

	equivalent, and even though the logic $\IKfour$, i.e., the
	intuitionistic version of the modal logic $\Kfour$, contains both
	axioms, they can also be added independently to the logic $\IK$ (an
	intuitionistic version of $\K$). The proof theory of these
	new logics has not been studied before; no existing labelled (or
	label-free) proof system can handle them, even though the

@@

	\;,

@@

	respectively, have already been studied
	in~\cite{plotkin:stirling:86}.

@@

	$\le$-relation) of the frame is visible in an ordinary labelled
	sequent, we only have that a sequent $\Gamma$ is provable if and
	only if is satisfied in all \emph{graph-consistent}\footnote{This

@@

	models, 

@@

	as already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
	and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).

@@

form $\futs xy$ and of the form $\accs xy$. Consequently, we can easily translate the frame

@@

of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

@@

\review{The rules of their calculus are identical (except for the ones that relate the two accessibility relations) to the rules for the intuitionistic-alethic fragment	of the calculus for intuitionistic bimodal logic in this publication:
  • Maffezioli, P., Naibo, A. \& Negri, S., The Church-Fitch knowability paradox in the light of structural proof theory. Synthese, vol. 190, 2677–2716, 2013.}
\lutz{done in previous paragraph? can this be removed: }
%\old{
%Maffezioli, Naibo and Negri have considered in~\cite{maffezioli:etal:synthese13} a labelled system for intuitionistic bimodal epistemic logic which is multi-conclusion.}
%%
%\marianela{
%	Our system uses the same principle, both labelled and multi-conclusion sequents, but we use in a more general setting and extend it to a framework for many intuitionistic modal logics.
%	%
%	This eliminates the undesired discrepancy as, consequently, every rule in our system is invertible, i.e.~we never delete information in a bottom-up proof search.  Furthermore, we present in this contribution an internal cut elimination proof.}

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics.

>>>>

Standard labelled sequent calculi for logics with
standard Kripke semantics attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally use
\emph{relational atoms} of the form $\accs xy$ where 
the binary relation symbol $\rel$
%These calculi work best for logics with
%standard Kripke semantics, as in this case the relation $\rel$ 
is used to encode the accessibility relation in the Kripke models, such that
frame conditions corresponding to the desired logic can be directly
encoded as inference rules.
%\todo{Reviewer 1 complained about the ``best''}
Prominent examples are classical modal
logics and intuitionistic propositional logic, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be translated into the
inference rule
\begin{equation}
\label{eq:Rtrans}
\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas~\cite{negri:vonplato:01}.

@@

logics} whose Kripke semantics is based on \emph{birelational}

@@

	\ax{4^{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}

@@

	\rn{4^{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
%	\quadcm

@@

	equivalent, and even though the logic $\IKfour$ 
	(intuitionistic version of modal logic $\Kfour$) contains both
	axioms, they can also be added independently to the logic $\IK$ 
	(intuitionistic version of $\K$). The proof theory of these
	distinct logics has not been discussed before, but their

@@

	\;

@@

	respectively, have been studied in~\cite{plotkin:stirling:86}.

@@

	$\le$-relation) is visible in an ordinary labelled
	sequent, we only have that a formula is provable in an extension of $\IK$ beyond paths axioms 
	if and only if is satisfied in all \emph{graph-consistent}\footnote{This

@@

	models satisfying the corresponding conditions, 

@@

	as observed by Simpson
	and considered as an inelegant solution~\cite{simpson:phd}. % (see also \cite{mar:str:tableaux17}).

@@

form $\futs xy$ and of the form $\accs xy$. 
Consequently, following the same ``axiom-to-rule'' method as Simpson, we can translate the frame

@@

of logics that had been treated proof-theoretically before.  
Furthermore, the relation
between syntax and semantics can be recovered as one would expect: A formula is
provable in this system if and only if it is valid in all models satisfying the corresponding frame condition.

@@

\review{The rules of their calculus are identical (except for the ones that relate the two accessibility relations) to the rules for the intuitionistic-alethic fragment	of the calculus for intuitionistic bimodal logic in this publication:
  • Maffezioli, P., Naibo, A. \& Negri, S., The Church-Fitch knowability paradox in the light of structural proof theory. Synthese, vol. 190, 2677–2716, 2013.}
\lutz{done in previous paragraph?}
%\old{
%Maffezioli, Naibo and Negri have considered in~\cite{maffezioli:etal:synthese13} a labelled system for intuitionistic bimodal epistemic logic which is multi-conclusion.}
%%
%\marianela{
%	Our system uses the same principle, both labelled and multi-conclusion sequents, but we use in a more general setting and extend it to a framework for many intuitionistic modal logics.
%	%
%	This eliminates the undesired discrepancy as, consequently, every rule in our system is invertible, i.e.~we never delete information in a bottom-up proof search.  Furthermore, we present in this contribution an internal cut elimination proof.}

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:completeness} and~\ref{sec:soundness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture intuitionistic Scott-Lemmon modal logics.

### 2020-09-29 FullyLabelled/JLC/main.tex
progress on section 7

\title{A fully labelled proof system for intuitionistic modal logics}
\author[1]{Sonia Marin} 
\author[2,3]{Marianela Morales}
\author[3,2]{Lutz Stra{\ss}burger}
\affil[1]{University College London}
\affil[2]{Laboratoire d'informatique d'\'Ecole polytechnique}
\affil[3]{Inria Saclay}

@@

 \begin{abstract}
   \review{Reconsider de abstract. It reads very “operational” and does not clearly state the contributions of the paper.}
   \lutz{Done?}
   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols appearing in sequents: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than existing labelled systems.
   %%
   We show soundness and completeness of our system, together with an internal cut elimination proof.
   %%
   Finally, we show how the system can be extended with arbitrary intuitionistic Scott-Lemmon axioms.
 \end{abstract}

>>>>

\title{A fully labelled proof system for intuitionistic modal logics}
\author[1]{Sonia Marin} 
\author[2,3]{Marianela Morales}
\author[3,2]{Lutz Stra{\ss}burger}
\affil[1]{University College London}
\affil[2]{Laboratoire d'Informatique de l'\'Ecole Polytechnique}
\affil[3]{Inria Saclay}

@@

 \begin{abstract}
   \review{Reconsider abstract. It reads very “operational” and does not clearly state the contributions of the paper.}
   \lutz{Done?}
   \sonia{Added two setences at the beginning and modified it a bit.}
   Labelled proof theory has been famously successful for modal logics by mimicking their relational semantics within deductive systems. Simpson in particular designed a framework to study a variety of intuitionistic modal logics integrating a binary relation symbol in the syntax.
   %%
   In this paper, we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols appearing in sequents: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. 
   %%
   As a consequence it can be extended with arbitrary intuitionistic Scott-Lemmon axioms. 
   %%
   We show soundness and completeness, together with an internal cut elimination proof, encompassing a wider array of intuitionistic modal logics than any existing labelled systems.
 \end{abstract}
